Assessing Amazon turker and automated machine forecasts in the Hybrid Forecasting Competition

### 2018-11-30 paper/paper.Rmd
paper writing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library("tidyverse")
```

```{r load-data}
dropbox_path <- "~/Dropbox/Work/asia-polmeth-2019"
in_dir  <- file.path(dropbox_path, "input-data")
out_dir <- file.path(dropbox_path, "output")

fcasts <- read_rds(file.path(out_dir, "user_forecasts_w_features.rds")) %>%
  mutate(date = as.Date(date),
         x = as.numeric(date) - min(as.numeric(date)),
         sees_chart = sees_chart==1,
         sees_model = sees_model==1) 
```

@@

### Turker

Due to lower than expected activity levels, turker forecasters started to be provided several weeks intot he first trial period. Given activity levels at that time, a decision was made to assign turkers to either condition A or C, but not B. This mean that the group whose treatment was to only see charts (B) consisted only of volunteers, while the other two groups contained mixed populations of volunteers and turkers, thus adding a confounding factor for assessing the impact of charts and machine models.

### Time series data was not available for all questions

Additionally, chartable data was only available for about a third of the questions. The chart and model treatments thus are really part of a nested three layer set of factors: whether a question had chartable data, then whether a chart was displayed to a user, then whether a machine forecast was also displayed to a user. Whether a question had associated data depended very much on the type of question, ranging from questions like "What will be the gold price on [date]" where there is a clearly relevant time series that is easy to acquire and update, to "How many battle deaths will ACLED report in [country] in [date]?" where data are available but have to be transformed to a relevant series, to, on the other end, questions like "Who will win the World Cup?" where it is exceedingly difficult to acquire relevant time series data. Thus it is also possible that questions with chartable and without chartable data were systemtically different in their difficutly and average acccuracy levels.

The effects of the two randomly assigned treatments, seeing a chart and seeing a model, are thus confounded by two other non-ranom factors: whether a forecaster was a volunteer or turker, and whether a question had chartable data or not. This effectively gives us an experiment design like in Table XXXX.

| Condition | Is Turker? | Chartable data | Sees chart? | Sees models? |
|-----------|------------|----------------|-------------|--------------|
| A.1 |   |
| A.2 | X |
| A.3 |   | X |
| A.4 | X | X |
| B.1 |   |   |
| B.2 |   | X | X |
| C.1 |   |   |
| C.2 | X |   |
| C.3 |   | X | X | X |
| C.4 | X | X | X | X |

@@

### Varying quality of data and machine forecasts over time

To complicate matters further, the quality of the data acquisition and machine forecasting modules changed over time during the RCT. Once a broader set of questions started flowing into the platform when the RCT started, a range of issues affecting both data quality and machine forecast quality arose. Examples include:

- Incorrect aggregation of event data like ACLED and ICEWS that produced incorrect monthly counts of events or casualties.
- Several questions were about monthly oil production rates for various countries. The resolution data are in PDF tables in monthly OPEC oil market reports. As this is hard to automatically scrape, an alternative source was used, but the values in this source do not always match the OPEC values. Nor, for that matter, do OPEC values always match in successive monthly reports. 
- Parsing failures that led the forecaster to produce non-sensical forecasts like negative values for count time series, or complete failures, e.g. when one question had a date like "April (04) 2018" instead of the expected "April 2018". 
- A limitation in the way Elastic Search aggregates non-standard fixed length time periods, e.g. periods of 40 days, required that for some questions data aggregation had to occur in the forecasting module, as opposed to data platform module, unlike planned.

Most of these issues and bugs were fixed during the course of the RCT, and thus the quality of both the data and machine forecasts should have improved over time.

@@

## Basic findings

### Brier over time

Make point here that Brier generaly decreased over time? Maybe questions easier, maybe forecasters learn over time? Just establish that all analysis from here is for May 2nd and on only.

### Other

Machine forecasts were not more accurate than human forecasts.

|Forecaster |Question         | A: no chart| B: chart only| C: chart and model|
|:----------|:----------------|-----------:|-------------:|------------------:|
|Volunteer  |IFP without data |        0.15|          0.09|               0.12|
|Volunteer  |IFP had data     |        0.25|          0.22|               0.28|
|Turker     |IFP without data |        0.15|            NA|               0.14|
|Turker     |IFP had data     |        0.28|            NA|               0.27|

Table: MMDB by forecaster, data availability, and condition.

The low MMDB for condition B forecasters appears to be primarily driven by good performance on questions that *did not* have charts (keeping in mind that 2/3 of questions would not have had a chart).

```{r tbl-accuracy-by-condition}
read_csv("../output/tables/brier-by-conditions.csv") %>%
  mutate(Condition = paste0(Condition, ".", c(1:4, 1:2, 1:4))) %>%
  knitr::kable(digits = 2)
```

Table: Mean Brier scores by condition

## Turker performance compared to volunteer forecasters

## Machine forecast impact

![Average mean daily Brier score (MDB) by forecaster type over time. Both volunteers and turkers improved over time.\label{fig:volunteers-only}](../output/figures/volunteers-only-accuracy-by-chartability-over-time.png)

Figure \ref{fig:volunteers-only} looks only at volunteer forecasters in conditions B and C, showing again mean daily Brier scores over time. The right chart shows that forecasters who saw both a chart and model had slighly worse performance throughout the RCT compared to condition B forecasters who only saw the chart. This matches the hypothesis that seeing the machine forecasts hurt performance. However, if we look at the chart on the left, which shows performance for questions for which no data was available, i.e. where both condition B and C forecasters saw neither a chart nor a model, condition C performance was also worse up until sometime in late July.

# Discussion/conclusion

Based on the quick check I did during the site visit, probably something like this:

- the machine forecasts were not accurate enough, although sans data and code issues they may have been 
- Turkers don't forecast as well as volunteers
- seeing the machine forecasts had a slight detrimental impact on accuracy, although maybe this changed as the forecasts became more accurate later in the RCT
- seeing the charts improved performance, but I don't think it is clear why. Seeing the chart for instances where the data are hard to acquire (e.g. some data transformation are needed) did not boost accuracy more than in other series.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

@@

Comparison of groups A and B should have shown the effect of seeing a time series chart, and of groups B and C for the effect of seeing a model forecast. In practice, there were issues that complicated the effective design or forecaster groupings and treatments:

- **Amazon Mechanical Turk forecasters**. Due to lower than expected activity levels, turker forecasters started to be provided several weeks intot he first trial period. Given activity levels at that time, a decision was made to assign turkers to either condition A or C, but not B. This mean that the group whose treatment was to only see charts (B) consisted only of volunteers, while the other two groups contained mixed populations of volunteers and turkers, thus adding a confounding factor for assessing the impact of charts and machine models. 
- **Gaps in data and machine coverage**. Only about 1/3 of IFPs had time series data available, meaning that within each condition group, most questions did not have a chart nor model. There was also a small number of instances where data was available, but a machine forecast was not. One example were questions related to FluNet influenza case counts, where a change in the data source broke the ability to update chart data, which would have required models to forecast over excessive time horizons. Since the availability of data was related to the type of question, it is not possible to rule out the possibility that questions with data were systematically different in their difficulty from questions without data. 
- **Changes in the data and machine forecasting platforms over time**. Both the data platform and the machine forecasting system suffered from various bugs and related issues, especially during the earlier portions of RCT-A. These problems in some cases resulted in incorrectly aggregated or otherwise inaccurate data, insufficient updating which led to data in the platform falling behind source availability and thus requiring forecasts over longer time periods than neccessary, and bugs in the machine forecaster that led to no or bad forecasts. The quality of the data and machine forecasts displayed to some users thus varied over time and IFPs, in ways that are difficult to reconstruct in retrospective.

In respect to the first two problems, which we can quantify, the effective treatment groups were more complicated and are showin in Table XX.

| Group|Condition          |Forecaster |IFP_Group       |Sees chart? |Sees model? |
|-----:|:------------------|:----------|:---------------|:-----------|:-----------|
|     1|A: no chart        |Turker     |No TS data      |            |            |
|     2|A: no chart        |Turker     |Chart only      |            |            |
|     3|A: no chart        |Turker     |Chart and model |            |            |
|     4|A: no chart        |Volunteer  |No TS data      |            |            |
|     5|A: no chart        |Volunteer  |Chart only      |            |            |
|     6|A: no chart        |Volunteer  |Chart and model |            |            |
|     7|B: chart only      |Volunteer  |No TS data      |            |            |
|     8|B: chart only      |Volunteer  |Chart only      |X           |            |
|     9|B: chart only      |Volunteer  |Chart and model |X           |            |
|    10|C: chart and model |Turker     |No TS data      |            |            |
|    11|C: chart and model |Turker     |Chart only      |X           |            |
|    12|C: chart and model |Turker     |Chart and model |X           |X           |
|    13|C: chart and model |Volunteer  |No TS data      |            |            |
|    14|C: chart and model |Volunteer  |Chart only      |X           |            |
|    15|C: chart and model |Volunteer  |Chart and model |X           |X           |
|    16|Machine            |Machine    |Chart and model |            |            |

@@

@@

|Forecaster | avg_Brier|     n|
|:----------|---------:|-----:|
|Machine    |     0.386|  1975|
|Turker     |     0.433| 39140|
|Volunteer  |     0.322|  7816|

Table: Average Brier score by forecaster group.

Volunteers who had only charts did the best

|Forecaster | A: no chart| B: chart only| C: chart and model| Machine|
|:----------|-----------:|-------------:|------------------:|-------:|
|Machine    |          NA|            NA|                 NA|   0.386|
|Turker     |       0.436|            NA|              0.432|      NA|
|Volunteer  |       0.297|          0.25|              0.372|      NA|

Table: Average Brier score by forecaster and condition groups.

Table XX summarizes the average Brier scores for all 16 treatment/forecaster/IFP groups in Table XX. As it is quite unwieldy, we include it for reference and will discuss specific insights in more details below.

| Group|Condition          |Forecaster |IFP_Group       | avg_Brier| sd_Brier|     n|
|-----:|:------------------|:----------|:---------------|---------:|--------:|-----:|
|     1|A: no chart        |Turker     |No TS data      |      0.45|     0.50|  6690|
|     2|A: no chart        |Turker     |Chart only      |      0.55|     0.56|   304|
|     3|A: no chart        |Turker     |Chart and model |      0.39|     0.40|  2907|
|     4|A: no chart        |Volunteer  |No TS data      |      0.31|     0.54|   510|
|     5|A: no chart        |Volunteer  |Chart only      |      0.36|     0.46|    18|
|     6|A: no chart        |Volunteer  |Chart and model |      0.28|     0.34|   274|
|     7|B: chart only      |Volunteer  |No TS data      |      0.26|     0.48|  1702|
|     8|B: chart only      |Volunteer  |Chart only      |      0.42|     0.42|    57|
|     9|B: chart only      |Volunteer  |Chart and model |      0.23|     0.31|   973|
|    10|C: chart and model |Turker     |No TS data      |      0.45|     0.50| 19502|
|    11|C: chart and model |Turker     |Chart only      |      0.56|     0.52|   990|
|    12|C: chart and model |Turker     |Chart and model |      0.38|     0.40|  8747|
|    13|C: chart and model |Volunteer  |No TS data      |      0.38|     0.60|  3090|
|    14|C: chart and model |Volunteer  |Chart only      |      0.59|     0.57|    73|
|    15|C: chart and model |Volunteer  |Chart and model |      0.32|     0.40|  1119|
|    16|Machine            |Machine    |Chart and model |      0.39|     0.36|  1975|

Table: Average Brier scores for all 15 distinct human IFP/treatment groups, as well as the machine forecasts.

## Turkers generally did worse

Turkers overall had worse forecast accuracy than either the machine models or volunteers. To rule a spurious relationship, we can compare the accuracy in directly comparable pairs, e.g. rows 1 and 4 in Table XX consists of the average Brier scores for questions without TS data and in condition A treatment group for turkers and volunteers respectively, and show a difference of 0.14.

Figure \ref{fig:turkers} shows the results of pairwise comparisons of the average Brier scores for all groups relative to Turker forecasters. Each row in the plot corresponds to a linear model for forecasters from the corresponding condition and IFP group, and the pointranges show coefficient estimates and 95\% confidence intervals for Turkers, Machine, and Volunteer forecasters. The Turker values correspond to the average Brier for them in this grouping, while the Machine and Volunteer estimates indicate the *relative* performance. I.e. for them, values below zero indicate an improvment. Note that the x-scale is reversed, so that better is further to the right.

\begin{figure}
\label{fig:turkers}
\caption{Pairwise comparisons of turker versus machine and volunteer forecast accuracy. Each row shows estimates from a linear regression model with turkers as reference group, for different groupings of the experimental conditions and IFP data/model availability. The turker estimates (red dots) correspond to the average Brier for turker forecasters, while the other estimates correspond to the relative difference from the turker average for other forecasters. Points further to the right correspond to lower Brier scores, i.e. better peformance.}
\includegraphics{../output/figures/pairwise-comparisons-turker.png}
\end{figure}

In most groups volunteers have an advantage sufficient to generate p-values below 0.05, although the effects are minute in comparison to general between-forecaster differences: all models have very small adjusted R^2 values, below 0.01.

## Chart and model effects

Teasing out the chart and model effects is a bit harder through pairwise comparison.

Add in the LM and LMER model results here.

```
Coefficients:
                          Estimate Std. Error t value Pr(>|t|)    
(Intercept)               0.447126   0.002887 154.862  < 2e-16 ***
ForecasterVolunteer      -0.104321   0.006208 -16.805  < 2e-16 ***
IFP_GroupChart only       0.142377   0.017834   7.984 1.45e-15 ***
IFP_GroupChart and model -0.063318   0.008648  -7.322 2.48e-13 ***
sees_chartTRUE           -0.038067   0.015751  -2.417   0.0157 *  
sees_modelTRUE            0.042337   0.015251   2.776   0.0055 ** 
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.4776 on 46950 degrees of freedom
Multiple R-squared:  0.01339,	Adjusted R-squared:  0.01328 
F-statistic: 127.4 on 5 and 46950 DF,  p-value: < 2.2e-16
```

```
Linear mixed model fit by REML ['lmerMod']
Formula: brier ~ Forecaster + IFP_Group + sees_chart + sees_model + (1 |      ifp_id)
   Data: common_period[common_period$Forecaster != "Machine", ]

REML criterion at convergence: 50272.9

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-2.9464 -0.5432 -0.2071  0.3053  4.9013

Random effects:
 Groups   Name        Variance Std.Dev.
 ifp_id   (Intercept) 0.06386  0.2527  
 Residual             0.16808  0.4100  
Number of obs: 46956, groups:  ifp_id, 156

Fixed effects:
                          Estimate Std. Error t value
(Intercept)               0.438521   0.025392  17.270
ForecasterVolunteer      -0.077777   0.005519 -14.091
IFP_GroupChart only       0.160131   0.107529   1.489
IFP_GroupChart and model -0.061051   0.044711  -1.365
sees_chartTRUE           -0.021495   0.013689  -1.570
sees_modelTRUE            0.026931   0.013280   2.028

Correlation of Fixed Effects:
            (Intr) FrcstV IFP_Go IFP_am ss_cTRUE
FrcstrVlntr -0.042                              
IFP_GrpChro -0.237  0.029                       
IFP_GrpCham -0.567  0.005  0.140                
ss_chrtTRUE  0.011 -0.266 -0.099 -0.066         
ses_mdlTRUE -0.012  0.284  0.081 -0.016 -0.816  
```

Relatively consistent resulst. First, turkers have worse performance. Second, seeing a chart somewhat helps, seeing a model somewhat hurts.

## Models generally were disappointing, but did they do well in certain areas?

This is now only looking at IFPs for which a machine forecast was available.

Put the model based version with stars here instead of this:

|Data_source      | N_IFPs| Machine| Turker| Volunteer|
|:----------------|------:|-------:|------:|---------:|
|ACLED 2op        |      7|   0.611|  0.533|     0.396|
|ACLED 5op        |      4|   0.371|  0.554|     0.790|
|earthquakes      |      2|   0.285|  0.207|     0.199|
|eia              |      5|   0.184|  0.204|     0.145|
|fao-fpi          |      3|   0.386|  0.220|     0.121|
|fred             |      2|   0.580|  0.488|     0.290|
|gold             |      3|   0.184|  0.274|     0.171|
|imf              |      1|   0.222|  0.252|     0.325|
|nigeria-security |      2|   0.355|  0.504|     0.648|
|oecd             |      7|   0.430|  0.534|     0.356|
|opec             |      5|   0.283|  0.222|     0.209|
|other            |      2|   0.416|  0.574|     0.188|
|privacyrights    |      2|   0.231|  0.321|     0.517|
|stocks           |      4|   0.204|  0.204|     0.223|

Models did relatively well with ACLED 5op, IMF, NST, privacyrights. Poorly on ACLED 2op, FAO, FRED, other?.

## Why did condition B volunteers do so well?

Condition B volunteers had the best overall performance out of the 16 groups shown in Table XX. One immediate and plausible story to explain why condition B forecasters did better than both A and C, even after controlling for the impact of Turker forecasters in the other groups, goes like this: we know the machine forecasts were disappointing and less accurate in general. Users who thus saw the chart and machine forecasts were adversely pulled to forecast in line with the machine prediction. On the other than, users who only saw the chart had a good baseline and were able to somehow accurately extrapolate from there.

There are several inconsistencies in the data with this story. First, condition B forecasters did not only do better than other groups on questions that had data and a chart, but also on questions that did not have a chart.

**evidence here**

Maybe there is a "spillover" effect, where seeing the charts somehow made users just generally better forecasters.

Do we see this in condition C for questions that didn't have a model? No. Condition C Turkers forecasting on a question without a chart did as poorly as condition A Turkers on the same set of questions (rows 1,2,10,11). Condition C volunteers on questions without a machine forecast did *worse* than both condition B and A volunteers on the same question sets.

**evidence here**

Another piece of evidence against the spillover effect is that accuracy is actually higher among users who only forecasted a small number of times, which suggests that learning did not take place.

**evidence here**

Not only were condition B volunteers uncannily accurate on questions even without a chart, they were uncannily accurate on questions where we know the displayed data were wrong.

**evidence here**

One possibility that we should not discard is that we are digging in noise. All of the effects---for turkers versus other forecasts, for the disappointing performance of the machine forecasts, for conditions or different groupings of IFPs---pale in comparison to differences in accuracy between individual forecasters and between forecasts for different IFPs.

**maybe this is a good time to talk about the superactive forecasters and see whether excluding them changes results**

# Conclusion

Are we digging in noise?

Many of the questions involve monthly time series but are only open for periods spanning a single or few months. Some of the advantage of the human forecasters may thus have been due to operating on a sub-monthly time scale.

We also know that model performance suffered from issues related to automation. Were we able to run the RCT-A questions with the current system and with a data platform that optimally ingests data sources on the bleeding edge of their own updates, our forecasts would have been ...

**check with updated results from test bed what the performance gain would have been, i think the site visit stats were not accurate anymore**
